{\Large \bf Analysing time course microarray data using Bioconductor: a case study using yeast2 Affymetrix arrays}

### 2012-11-20 paper.Rnw
Minor code formatting changes

<<eval=FALSE>>=
url='http://bioconductor.org/biocLite.R'
source(url)
biocLite()
@

@@

<<eval=TRUE,message=FALSE,warning=FALSE>>=
#Read in the mask file
s_cer = read.table('s_cerevisiae.msk', skip=2, 
stringsAsFactors=FALSE)
source('ExtractIDs.R')
c_df = ExtractIDs(s_cer[ ,1])
@
\noindent We also need to restrict the view of \cc{yeast.raw} to the $x-$ and $y-$coordinates of the \textit{S. cerevisiae} probesets in the \cc{cdf} environment by using
<<eval=TRUE,message=FALSE>>=
#Get the raw dataset for S. cerevisiae only
library(affy)
library(yeast2probe)
source('RemoveProbes.R')
cleancdf = cleancdfname(yeast.raw@cdfName)
RemoveProbes(s_cer[ ,1], cleancdf, 'yeast2probe') 
@
Note that the commands in \cc{RemoveProbes.R} are listed in Appendix A. Thus the attributes of \cc{yeast.raw}, obtained via
\cc{print(yeast.raw)}, are now
{\small\begin{verbatim}
    AffyBatch object
    size of arrays = 496x496 features (3167 kb)
    cdf = Yeast_2 (5900 affyids)
    number of samples = 30
    number of genes = 5900
    annotation = yeast2
\end{verbatim}}
\noindent and the number of genes (actually probesets here) is 5,900 now that the \textit{S. pombe} probesets have been removed.

@@

Before any formal statistical analysis, it is important to check for
data quality. Initially, we might examine the perfect and mismatch
probe-level data to detect anomalies. Images of the first five arrays can be obtained using
<<s_fig1,eval=FALSE,tidy=TRUE>>=
for(i in 1:5){
    plot_title = paste('Strain: ', exp_fac$strain[i], 
        'Time: ', exp_fac$tps[i])
        d = exp_fac$data_order[i]
    image(yeast.raw[,d], main=plot_title) 
}
@

@@

<<eval=TRUE,cache=TRUE,results=FALSE,message=FALSE>>=
yeast.rma = rma(yeast.raw)
yeast.matrix = exprs(yeast.rma)[ ,exp_fac$data_order]
colnames(yeast.matrix) = paste(exp_fac$strain, exp_fac$tps, sep='')
exp_fac$data_order = 1:30
@

@@

<<eval=TRUE,cache=TRUE>>=
#Obtain the maximum fold change but keep the sign 
maxfoldchange = function(foldchange)
foldchange[which.max(abs(foldchange))]  
difference = apply(eb$coeff, 1, maxfoldchange)
pvalue = eb$F.p.value
lodd = -log10(pvalue)

@@

<<F5,eval=TRUE,dev="pdf",fig.ext='pdf',fig.pos="!t", cache=TRUE, fig.width=6.5,fig.height=4,out.width="0.8\\textwidth",fig.cap="Eight clusters obtained from the \texttt{Mfuzz} package.",message=FALSE>>=
library(Mfuzz)
tmp_expr = new('ExpressionSet', exprs=m)
cl = mfuzz(tmp_expr, c=8, m=1.25)
mfuzz.plot(tmp_expr,cl=cl, mfrow=c(2, 4), new.window = FALSE)
@
Of course, it is usually not clear how many clusters there are (or
should be) within a dataset and so the sensitivity of conclusions to
the choice of number of clusters ($c$) should always be
investigated. For example, if $c$ is chosen to be too large then some
clusters will appear sparse and this might suggest choosing a smaller
value of~$c$. Figure~\ref{F5} shows the profiles of the
eight clusters obtained from the \cc{Mfuzz} package. The probes 
present within each cluster can be found by using

>>>>

<<eval=FALSE>>=
url = 'http://bioconductor.org/biocLite.R'
source(url)
biocLite()
@

@@

<<eval=TRUE,message=FALSE,warning=FALSE>>=
#Read in the mask file
s_cer = read.table('s_cerevisiae.msk', skip=2, 
stringsAsFactors=FALSE)
source('ExtractIDs.R')
c_df = ExtractIDs(s_cer[ ,1])
@
\noindent We also need to restrict the view of \cc{yeast.raw} to the $x-$ and $y-$coordinates of the \textit{S. cerevisiae} probesets in the \cc{cdf} environment by using
<<eval=TRUE,message=FALSE>>=
#Get the raw dataset for S. cerevisiae only
library(affy)
library(yeast2probe)
source('RemoveProbes.R')
cleancdf = cleancdfname(yeast.raw@cdfName)
RemoveProbes(s_cer[, 1], cleancdf, 'yeast2probe') 
@
Note that the commands in \cc{RemoveProbes.R} are listed in Appendix A. Thus the attributes of \cc{yeast.raw}, obtained via
\cc{print(yeast.raw)}, are now
{\small\begin{verbatim}
    AffyBatch object
    size of arrays = 496x496 features (3167 kb)
    cdf = Yeast_2 (5900 affyids)
    number of samples = 30
    number of genes = 5900
    annotation = yeast2
\end{verbatim}}
\noindent and the number of genes (actually probesets here) is 5,900 now that the \textit{S. pombe} probesets have been removed.

@@

Before any formal statistical analysis, it is important to check for
data quality. Initially, we might examine the perfect and mismatch
probe-level data to detect anomalies. Images of the first five arrays can be obtained using
<<s_fig1,eval=FALSE,tidy=TRUE>>=
for(i in 1:5) {
    plot_title = paste('Strain: ', exp_fac$strain[i], 
        'Time: ', exp_fac$tps[i])
        d = exp_fac$data_order[i]
    image(yeast.raw[,d], main=plot_title) 
}
@

@@

<<eval=TRUE,cache=TRUE,results=FALSE,message=FALSE>>=
yeast.rma = rma(yeast.raw)
yeast.matrix = exprs(yeast.rma)[, exp_fac$data_order]
colnames(yeast.matrix) = paste0(exp_fac$strain, exp_fac$tps)
exp_fac$data_order = 1:30
@

@@

<<eval=TRUE,cache=TRUE>>=
#Obtain the maximum fold change but keep the sign 
maxfoldchange = function(foldchange){
    foldchange[which.max(abs(foldchange))]
}

difference = apply(eb$coeff, 1, maxfoldchange)
pvalue = eb$F.p.value
lodd = -log10(pvalue)

@@

<<F5,eval=TRUE,dev="pdf",fig.ext='pdf',fig.pos="!t", cache=TRUE, fig.width=6.5,fig.height=4,out.width="0.8\\textwidth",fig.cap="Eight clusters obtained from the \\texttt{Mfuzz} package.",message=FALSE>>=
library(Mfuzz)
tmp_expr = new('ExpressionSet', exprs=m)
cl = mfuzz(tmp_expr, c=8, m=1.25)
mfuzz.plot(tmp_expr,cl=cl, mfrow=c(2, 4), new.window = FALSE)
@
Of course, it is usually not clear how many clusters there are (or
should be) within a dataset and so the sensitivity of conclusions to
the choice of number of clusters ($c$) should always be
investigated. For example, if $c$ is chosen to be too large then some
clusters will appear sparse and this might suggest choosing a smaller
value of~$c$. Figure~\ref{F5} shows the profiles of the
eight clusters obtained from the \cc{Mfuzz} package. The probes 
present within each cluster can be found by using
